Nucleotide Diversity Loss on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-11-01 manuscript.tex
Update manuscript.tex

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which can significantly affect levels of genetic variability within the genome. On Y chromosomes with suppressed recombination, selection can remove genetically-linked neutral variation, resulting in reduced $N_{e}$ and the level of polymorphism on Y compared to X chromosomes or autosomes. However, non-selective factors including female biased sex ratios and high variance in male fitness can also reduce Y-linked $N_{e}$, making it difficult to infer the causes of low Y-chromosome diversity. Here, we investigate the factors affecting X- and Y-linked diversity during plant sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae). Strikingly, we find that average neutral polymorphism levels for genes on the Y chromosome are \textasciitilde 2.1\% the value of their homologues on the X chromosome, indicating a significant departure from neutrality and a chromosome-wide reduction of \textasciitilde 93\%. We demonstrate that the magnitude of this diversity loss is inconsistent with neutral models that account for the occurrence of female-biased sex ratios, or by reduced male $N_{e}$ caused by high variance in male fitness. Rather, using forward simulations and estimates of the distribution of fitness effects of deleterious mutations, we show that neutral polymorphism on the Y can be explained by the effects of linked purifying selection, although selective sweeps may have also contributed. Given the recent origin of \textit{R. hastatulus} sex chromosomes, our study suggests that the early stages of Y chromosome degeneration are characterized by strong selective interference among many linked sites rather than gene silencing followed by neutral genetic drift.
%abstract is already 270 words. Should be 250 (but might be ok). Just can't get longer.
\end{abstract}

@@

One fundamental difference between the X and Y chromosomes is that there are 1/3 as many Y-linked gene copies as X-linked ones in a diploid population. Because of this, the effective population size ($N_{e}$) of the Y chromosome is expected to be 1/4 that of the autosomes, whereas the $N_{e}$ for the X chromosome should be 3/4 that of autosomes (assuming an equal number of reproducing females and males). The lowered $N_{e}$ of the Y chromosome implies that delterious mutations are more likely to rise to high frequency because of gentic drift \citep{nei1981genetic} - an effect that can be exacerbated by selective interference among genetically linked sites (the "Hill-Robertson Effect"). These effects both predict a reduction in the level of neutral polymorphism maintained on the Y chromosome, which at equilibrium is proportional to the product of $N_{e}$ and the mutation rate, $\mu$ \citep{Kimura1984,charlesworth1987}.

In species with female-biased reproductive sex ratios or extensive male-male competiton, a high variance in male reproductive success could also reduce the $N_{e}$ of the Y chromosome \citep{caballero1995,charlesworth2001,laporte2002,pool2007,ellegren2009}, suggesting the need to distinguish the effects of variance in reproductive success from the effects selective interference just mentioned. Because high variance in male reproductive success is expected to reduce both Y-linked $N_{e}$ and and autosomal $N_{e}$ \citep{kimura1964number,nomura2002effective}, evidence for this effect can be obtained by examining levels of neutral variability on X and Y chromosomes relative to values for the autosomes. In particular, high variance in male reproductive success predicts a reduction in the Y/A diversity ratio, but an increase in X/A. Based on this comparison, studies of the X/A ratio in humans have suggested a historical excess of breeding females over males resulting in an inflated X/A ratio (though there is considerable variation in estimates of X/A in humans, and other demographic scenarios have also been suggested \citep{hammer2010,bustamante2009}).
%as modeled by \citep{kimura1964number} and \citep{nomura2002effective}.

@@

Of particular relevance to the present study, \textit{R. hastatulus} populations have also been found to consistently exhibit female-biased sex ratios, with a mean sex ratio of $N_{f}/(N_{m}+N_{f})=0.6$ \citep{pickup2013influence}. Using genome-wide estimates of neutral polymorphism together with forward population genetic simulations, we test whether levels of genetic diversity on sex chromosomes and autosomes in this species are consistent with neutral evolutionary processes, including the joint effects of sex ratio bias and variance in reproductive success, or whether the removal of linked neutral variation caused by purifying selection has played a more significant role.

\section*{Materials and Methods}
\subsection*{Population Samples and Sex-Linked Genes}
We analyzed sex-linked and autosomal genes identified from Illumina RNA sequence data from 24 individuals (12 males and 12 females, with 1 male and 1 female from each of 12 populations; 6 populations of each sex chromosome race). Samples were collected in 2010 from throughout the native range of \textit{R. hastatulus} (locations in Table S1), and plants were grown in the glasshouse from seeds collected from open-pollinated females. We extracted RNA from leaf tissue using Spectrum Plant Total RNA kits (Sigma-Aldrich). The isolation of mRNA and cDNA synthesis was conducted according to standard Illumina RNAseq procedures, with sequencing conducted on two Illumina HiSeq lanes with 150-bp end reads at the Genome Quebec Innovation Center. Reads from these 24 samples were mapped to the \textit{R. hastatulus} reference transcriptome \citep{hough2014}, and raw sequences are available from the Gen-Bank Short Read Archive under accession no. SRP041588. Reads were mapped using the Burrows–Wheeler Aligner \citep{li2010fast}, followed by Stampy \citep{lunter2011stampy}. We used Picard tools (http://picard.sourceforge.net) to modify mapping output into the format required for the Genome Analysis Toolkit \citep{mckenna2010genome} variant calling software, and subsequently removed genes with low coverage (<10x) and Phred Quality Scores <20. The population samples analyzed here were previously reported in Hough et al. (2014), where they were used to validate the ascertainment of sex-linked genes identified through segregation analysis. Here, to consider sex linked genes that were identified in each of our sequenced population samples, we focused on the set of 460 X/Y genes for which the Y homolog was found in both the Texas and North Carolina races (i.e., X/Y genes where the Y copy was inferred to be on the $Y_{1}$ chromosome).

\subsection*{Autosomal Genes}
In evaluating evidence for nucleotide diversity differences between X and Y chromosomes, it is important to distinguish between reduced Y-linked diversity, and the possibility that X-linked diversity is elevated above the level predicted from a neutral model. To do this, we normalized sex-linked diversity estimates by autosomal diversity, and compared empirical X/A and Y/A nucleotide diversity ratios to those predicted from neutral models and from simulations (described below). Because the criteria for identifying autosomal loci in Hough et al. (2014) were based on the occurrence of four segregating SNPs per locus, this set of genes is probably higher in diversity than the average autosome. Therefore, for the present analysis we incorporated the broader set of all non-sex linked (putatively autosomal) genes from our transcriptome data. We also filtered this set to remove genes that may have been sex-linked but were not identified as such by Hough et al.'s conservative ascertainment criteria. In particular, we removed: (i) any genes in which there was evidence for at least one SNP with a sex-linked segregation pattern, (ii) any genes with SNPs showing fixed heterozygosity in males and fixed homozygosity in females, (iii) any genes with less than 10X coverage or greater than 100X coverage from independently obtained genomic coverage data (to filter out duplicate genes or those with highly repetitive sequences), and (iv) any genes containing SNPs with large (>0.4) allele frequency differences between males and females. Finally, we removed genes with fewer than 100 synonymous sites following this filtering to avoid biasing our results toward genes that may have been particularly short due to assembly problems. This filtering resulted in a set of 12,356 and 11,350 autosomal genes in the Texas and North Carolina races, respectively.

@@

\subsection*{Estimating nucleotide diversity on sex chromosomes and autosomes}
For each locus in our analysis, we calculated Watterson’s (1975) estimator of the population parameter $\theta=4N_{e}\mu$, where $N_{e}$ is the effective population size, and $\mu$ is the mutation rate \citep{watterson1975}, using a modified version of the Perl program Polymorphurama \citep{bachtrog2006}. To compare sex-linked and autosomal loci, we calculated the average value of $\theta$, weighted by the number of synonymous sites in each gene (\X Figure 2; Table 1). We obtained 95 percent confidence intervals for our estimates of the X/A and Y/A diversity ratios by bootstrapping by gene using the BCa method \citep{efron1994} implemented in the Boot package in R \citep{canty2012boot}, and calculating X/A and Y/A diversity on each iteration for 20000 replicates each. Bootstrapping was conducted on the final filtered set of 173 sex-linked, and 12355 autosomal genes from the Texas race, and separately for the 176 sex-linked and 11349 autosomal genes from the North Carolina race. Note that the lack of recombination on the Y chromosome implies that assumptions about independence across loci are likely violated, suggesting that the true uncertainty in our Y/A diversity estimate may be wider than implied by bootstrapping across genes. To address this issue, we also used a maximum likelihood approach, implemented in a modified version of the MLHKA software \citep{wright2004hka} to independently estimate a credibility interval for the Y/A ratio using coalescent theory (Figure S1). Because of the thousands of genes involved, a likelihood method incorporating divergence to control for heterogeneity in mutation rate is not feasible, since this would involve maximizing the likelihood estimate of the mutation rate for each locus independently.  Therefore, we applied a method that assumed no heterogeneity in mutation rate, no recombination between Y-linked genes, and free recombination between autosomal loci. Our model therefore had two parameters; theta for autosomal genes and the Y/A ratio. We varied both parameters across a grid of points and evaluated the likelihood, where the Y/A ratio varied from 0.001 to 1, and Watterson's theta on the autosomes varied from 0.001 to 0.01. We also tested whether estimates of diversity for the X chromosome calculated from phased sequences from females were consistent with estimates from phased sequences from males. As no significant difference was observed (\X Figure? Table?), we report only results from females.

@@

\subsection*{Simulations of purifying selection}
To test whether our observed level of Y-chromosome diversity could be explained by the effects of linked purifying selection, we followed the approach used in \citep{Wilsonsayres2014} and conducted forward-time simulations of a nonrecombining Y chromosome using the software SFSCODE \citep{hernandez2008flexible} and compared our empirical diversity estimates on the Y with those from simulations. This approach was preferred over using analytical predictions of background selection primarily because the equilibrium background selection model over-predicts the reduction in diversity when there are many linked sites under selection \citep{KaiserCharlesworth}, as is expected to be the case for large Y chromosomes that lack crossing over.

@@

\begin{table}[t!]
\centering
\caption{Estimates of neutral diversity by race on \textit{R. hastatulus} sex chromosomes and autosomes.}
\begin{tabular}{ccccccccc}
\textbf{} & \multicolumn{2}{c}{\textbf{Texas}} & \multicolumn{2}{c}{\textbf{South Carolina}} & \multicolumn{2}{c}{\textbf{Florida}} \\
chromosome & $\theta$ & $\theta/\theta_{A}$ & $\theta$ & $\theta/\theta_{A}$ & $\theta$ & $\theta/\theta_{A}$ \\
\midrule
A & 0.006 & 1 & 0.006 & 1 & 0.005 & 1 \\
X & 0.0047 & 0.85 & 0.002 & 0.33 & 0.0047 & 0.37 \\
Y & $\num{e-4}$ & 0.002 & $\num{e-4}$ & 0.002 & $\num{e-4}$ & 0.002 \\
\addlinespace

>>>>

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which can significantly affect levels of genetic variability within the genome. On Y chromosomes with suppressed recombination, selection can remove genetically-linked neutral variation, resulting in reduced $N_{e}$ and the level of polymorphism on Y compared to X chromosomes or autosomes. However, non-selective factors including female biased sex ratios and high variance in male fitness can also reduce Y-linked $N_{e}$, making it difficult to infer the causes of low Y-chromosome diversity. Here, we investigate the factors affecting X- and Y-linked diversity during plant sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae). Strikingly, we find that average neutral polymorphism levels for genes on the Y chromosome are \textasciitilde 2.1\% the value for autosomes, indicating a significant departure from neutrality. We demonstrate that the magnitude of this diversity loss is inconsistent with neutral models that account for the occurrence of female-biased sex ratios, or by reduced male $N_{e}$ caused by high variance in male fitness. Rather, using forward simulations and estimates of the distribution of fitness effects of deleterious mutations, we show that neutral polymorphism on the Y can be explained by the effects of linked purifying selection, although selective sweeps may have also contributed. Given the recent origin of \textit{R. hastatulus} sex chromosomes, our study suggests that the early stages of Y chromosome degeneration are characterized by strong selective interference among many linked sites rather than gene silencing followed by neutral genetic drift.
%abstract is already 270 words. Should be 250 (but might be ok). Just can't get longer.
\end{abstract}

@@

One fundamental difference between the X and Y chromosomes is that there are 1/3 as many Y-linked gene copies as X-linked ones in a diploid population. Because of this, the effective population size ($N_{e}$) of the Y chromosome is expected to be 1/4 that of the autosomes, whereas the $N_{e}$ for the X chromosome should be 3/4 that of autosomes (assuming an equal number of reproducing females and males). The lowered $N_{e}$ of the Y chromosome implies that deleterious mutations are more likely to rise to high frequency because of gentic drift \citep{nei1981genetic} - an effect that can be exacerbated by selective interference among genetically linked sites (the "Hill-Robertson Effect"). These effects both predict a reduction in the level of neutral polymorphism maintained on the Y chromosome, which at equilibrium is proportional to the product of $N_{e}$ and the mutation rate, $\mu$ \citep{Kimura1984,charlesworth1987}.

In species with female-biased reproductive sex ratios or extensive male-male competiton, a high variance in male reproductive success could also reduce the $N_{e}$ of the Y chromosome \citep{caballero1995,charlesworth2001,laporte2002,pool2007,ellegren2009}, suggesting the need to distinguish the effects of variance in reproductive success from the effects selective interference just mentioned. Because high variance in male reproductive success is expected to reduce both Y-linked $N_{e}$ and autosomal $N_{e}$ \citep{kimura1964number,nomura2002effective}, evidence for this effect can be obtained by examining levels of neutral variability on X and Y chromosomes relative to values for the autosomes. In particular, high variance in male reproductive success predicts a reduction in the Y/A diversity ratio, but an increase in X/A. Based on this comparison, studies of the X/A ratio in humans have suggested a historical excess of breeding females over males resulting in an inflated X/A ratio (though there is considerable variation in estimates of X/A in humans, and other demographic scenarios have also been suggested \citep{hammer2010,bustamante2009}).
%as modeled by \citep{kimura1964number} and \citep{nomura2002effective}.

@@

Of particular relevance to the present study, \textit{R. hastatulus} populations have also been found to consistently exhibit female-biased sex ratios, with a mean sex ratio of $N_{f}/(N_{m}+N_{f})=0.6$ \citep{pickup2013influence}. Using genome-wide estimates of neutral polymorphism together with forward population genetic simulations, we test whether levels of genetic diversity on sex chromosomes and autosomes in this species are consistent with neutral evolutionary processes, including the joint effects of sex ratio bias and variance in reproductive success, or whether the removal of linked neutral variation caused by purifying selection is also playing a significant role.

\section*{Materials and Methods}
\subsection*{Population Samples and Sex-Linked Genes}
We analyzed sex-linked and autosomal genes identified from Illumina RNA sequence data from 24 individuals (12 males and 12 females, with 1 male and 1 female from each of 12 populations; 6 populations of each sex chromosome race). Samples were collected in 2010 from throughout the native range of \textit{R. hastatulus} (locations in Table S1), and plants were grown in the glasshouse from seeds collected from open-pollinated females. We extracted RNA from leaf tissue using Spectrum Plant Total RNA kits (Sigma-Aldrich). The isolation of mRNA and cDNA synthesis was conducted according to standard Illumina RNAseq procedures, with sequencing conducted on two Illumina HiSeq lanes with 150-bp end reads at the Genome Quebec Innovation Center. Reads from these 24 samples were mapped to the \textit{R. hastatulus} reference transcriptome \citep{hough2014}, and raw sequences are available from the GenBank Short Read Archive under accession no. SRP041588. Reads were mapped using the Burrows–Wheeler Aligner \citep{li2010fast}, followed by Stampy \citep{lunter2011stampy}. We used Picard tools (http://picard.sourceforge.net) to modify mapping output into the format required for the Genome Analysis Toolkit \citep{mckenna2010genome} variant calling software, and subsequently removed genes with low coverage (<10x) and Phred Quality Scores <20. The population samples analyzed here were previously reported in Hough et al. (2014), where they were used to validate the ascertainment of sex-linked genes identified through segregation analysis. Here, to consider sex linked genes that were identified in each of our sequenced population samples, we focused on the set of 460 X/Y genes for which the Y homolog was found in both the Texas and North Carolina races (i.e., X/Y genes where the Y copy was inferred to be on the $Y_{1}$ chromosome).

\subsection*{Autosomal Genes}
In evaluating evidence for nucleotide diversity differences between X and Y chromosomes, it is important to distinguish between reduced Y-linked diversity, and the possibility that X-linked diversity is elevated above the level predicted from a neutral model. To do this, we normalized sex-linked diversity estimates by autosomal diversity, and compared empirical X/A and Y/A nucleotide diversity ratios to those predicted from neutral models and from simulations (described below). Because the criteria for identifying autosomal loci in Hough et al. (2014) were based on the occurrence of four segregating SNPs per locus, this set of genes is probably higher in diversity than the average autosomal gene. Therefore, for the present analysis we incorporated the broader set of all non-sex linked (putatively autosomal) genes from our transcriptome data. We also filtered this set to remove genes that may have been sex-linked but were not identified as such by Hough et al.'s conservative ascertainment criteria. In particular, we removed: (i) any genes in which there was evidence for at least one SNP with a sex-linked segregation pattern, (ii) any genes with SNPs showing fixed heterozygosity in males and fixed homozygosity in females, (iii) any genes with less than 10X coverage or greater than 100X coverage from independently obtained genomic coverage data (to filter out duplicate genes or those with highly repetitive sequences), and (iv) any genes containing SNPs with large (>0.4) allele frequency differences between males and females. Finally, we removed genes with fewer than 100 synonymous sites following this filtering to avoid biasing our results toward genes that may have been particularly short due to assembly problems. This filtering resulted in a set of 12,356 and 11,350 autosomal genes in the Texas and North Carolina races, respectively.

@@

\subsection*{Estimating nucleotide diversity on sex chromosomes and autosomes}
For each locus in our analysis, we calculated Watterson’s (1975) estimator of the population parameter $\theta=4N_{e}\mu$, where $N_{e}$ is the effective population size, and $\mu$ is the mutation rate \citep{watterson1975}, using a modified version of the Perl program Polymorphurama \citep{bachtrog2006}. To compare sex-linked and autosomal loci, we calculated the average value of $\theta$, weighted by the number of synonymous sites in each gene (\X Figure 2; Table 1). We obtained 95 percent confidence intervals for our estimates of the X/A and Y/A diversity ratios by bootstrapping by gene using the BCa method \citep{efron1994} implemented in the Boot package in R \citep{canty2012boot}, and calculating X/A and Y/A diversity on each iteration for 20000 replicates each. Bootstrapping was conducted on the final filtered set of 173 sex-linked, and 12355 autosomal genes from the Texas race, and separately for the 176 sex-linked and 11349 autosomal genes from the North Carolina race. Note that the lack of recombination on the Y chromosome implies that assumptions about independence across loci are violated, suggesting that the true uncertainty in our Y/A diversity estimate may be wider than implied by bootstrapping across genes. To address this issue, we also used a maximum likelihood approach, implemented in a modified version of the MLHKA software \citep{wright2004hka} to independently estimate a credibility interval for the Y/A ratio using coalescent theory (Figure S1). Because of the thousands of genes involved, a likelihood method incorporating divergence to control for heterogeneity in mutation rate is not feasible, since this involves maximizing the likelihood estimate of the mutation rate for each locus independently.  Therefore, we applied a method that assumed no heterogeneity in mutation rate, no recombination between Y-linked genes, and free recombination between autosomal loci. Our model therefore had two parameters; theta for autosomal genes and the Y/A ratio. We varied both parameters across a grid of points and evaluated the likelihood, where the Y/A ratio varied from 0.001 to 1, and Watterson's theta on the autosomes varied from 0.001 to 0.01. We also tested whether estimates of diversity for the X chromosome calculated from phased sequences from females were consistent with estimates from phased sequences from males. As no significant difference was observed (\X Figure? Table?), we report only results from females.

@@

\subsection*{Simulations of purifying selection}
To test whether our observed level of Y-chromosome diversity could be explained by the effects of linked purifying selection, we followed the approach used in \citep{Wilsonsayres2014} and conducted forward-time simulations of a nonrecombining Y chromosome using the software SFSCODE \citep{hernandez2008flexible} and compared our empirical diversity estimates on the Y with those from simulations. This approach was implemented rather than using analytical predictions of background selection primarily because the equilibrium background selection model over-predicts the reduction in diversity when there are many linked sites under selection \citep{KaiserCharlesworth}, as is expected to be the case for large Y chromosomes that lack crossing over.

@@

\begin{table}[t!]
\centering
\caption{Estimates of neutral diversity by race on \textit{R. hastatulus} sex chromosomes and autosomes.}
\begin{tabular}{ccccccccc}
\textbf{} & \multicolumn{2}{c}{\textbf{Texas}} & \multicolumn{2}{c}{\textbf{South Carolina}} & \multicolumn{2}{c}{\textbf{Florida}} \\
chromosome & $\theta$ & $\theta/\theta_{A}$ & $\theta$ & $\theta/\theta_{A}$ & $\theta$ & $\theta/\theta_{A}$ \\
\midrule
A & 0.006 & 1 & 0.006 & 1 & 0.005 & 1 \\
X & 0.0047 & 0.85 & 0.0019 & 0.33 & 0.0018 & 0.37 \\
Y & $\num{e-4}$ & 0.002 & $\num{e-4}$ & 0.002 & $\num{e-4}$ & 0.002 \\
\addlinespace
